Artificial ecosystem selection reveals relationships between microbiome composition and ecosystem function

### 2023-02-20 manuscript/manuscript.Rmd
edit readme, recreate makefile, simplify analysis scripts

@@

```{r initialize-packages}
library(tidyverse)
library(broom)
library(broman)
library(morris)
# library(phyloseq)
library(ggpubr)
library(knitr)
library(corncob)
# library(ggthemes)
library(captioner)
library(knitr)
library(ggtext)
figures <- captioner()
tables <- captioner(prefix = "Table")

@@

# CH4 models
load(file = paste0(der_dir, 'response_model.RData'))
load(file = paste0(der_dir, 'response_plot.RData'))
load(file = paste0(der_dir, 'herit_model.RData'))
load(file = paste0(der_dir, 'herit_plot.RData'))

# 16S models
load(file = paste0(der_dir, 'richness_models.RData'))
load(file = paste0(der_dir, 'phyloseq.RData'))
load(file = paste0(der_dir, 'beta_model.RData'))
load(file = paste0(der_dir, 'beta_fig.RData'))
load(file = paste0(der_dir, 'da_corncob.RData'))
load(file = paste0(der_dir, 'da_fig.RData'))
load(file = paste0(der_dir, 'dbrda.RData'))

@@

Microbiomes mediate important ecosystem functions, yet it has proven difficult
to determine the relationship between microbiome composition and the rate of
ecosystem functions. This challenge remains because it is difficult to
manipulate microbiome composition directly, we often cannot know a priori which
members of a microbiome influence the rate of an ecosystem function, and
microbiomes can covary strongly with other drivers of ecosystem function, such
as the environment. We propose that artificial ecosystem selection can overcome
these challenges. Artificial ecosystem selection involves creating replicate
ecosystems, selecting replicates with a desired ecosystem-level trait, and
using the microbiomes from the selected ecosystems to inoculate a new set of
ecosystems. We performed artificial ecosystem selection on replicate soil
microbiomes to select for increased soil methane oxidation rate. We observed a
strong response to selection with a `r myround(per_slope, 1)`\% increase in
methane oxidation rate per passage. In addition, we
estimated that `r myround((coefficients(herit_model)[2] +
coefficients(herit_model)[4])*100, 1)`\% of the variation in methane oxidation
rate in these soils can be attributed to microbiome variation. We also found
that selection did not enrich for known methane oxidizers; rather microbes that
responded to selection included members of `r length(da_fams)` families not
known to oxidize methane, such as *Fimbriimonadaceae*, *Cytophagaceae*, and
*Diplorickettsiaceae*. This result is in contrast to the typical assumption that
the rate of an ecosystem function is limited by the final step in the associated
microbial pathway. Our study demonstrates that variation in microbiome
composition can contribute to variation in the rate of ecosystem function and
that manipulating microbiome composition could be a viable strategy for managing
ecosystem functions.

@@

Microbiomes mediate a variety of important ecosystem functions. This suggests
that understanding variation in microbiome composition might be important for
predicting the rates of ecosystem functions 
[@conrad1996; @schimel1998; @crowther2019]. 
Many studies have attempted to document such a relationship
[@cavigelli2000; @balser2005; @waldrop2006; @strickland2009; @wagg2014; @wagg2019].
However, it is difficult to isolate the effect of variation in microbiome
composition from other drivers of variation in ecosystem function. For example,
microbiome manipulations often confound variation in diversity, composition, and
abundance. Likewise, comparative studies must control for abiotic variation
statistically to identify the direct effect of microbiome variation. This
approach relies on measuring the relevant abiotic variables and cannot account
for unmeasured environmental variation. Here, we overcome these limitations by
using a selection approach to estimate the degree to which an ecosystem function
varies with microbiome composition.

There have been two general categories of approaches that investigators have
used to estimate the degree to which an ecosystem function varies with
microbiome composition: comparative and manipulative. Comparative
studies sample natural variation in an ecosystem function across different
habitats and simultaneously measure variation in community composition.
Investigators can then correlate ecosystem function with aspects of community
composition while attempting to control for environmental variation. These
approaches have
documented important relationships between microbiomes and ecosystem
functions. For example, a meta-analysis of these studies observed a small but
significant contribution of the microbiome to variation in ecosystem
function after controlling for environmental variation [@graham2016]. In
addition, studies focusing on the correlation between the rate of an ecosystem
function and the abundance of an associated functional marker gene observe a
significant correlation, though this relationship is rare and contingent upon
both the function and the ecosystem sampled [@rocca2015]. However, comparative
studies come with some unique challenges and limitations. One issue is that
microbiome attributes tend to covary with the abiotic conditions within an
environment, and whether and how the investigator controls for these variables
can affect the conclusions of such a comparative study. In addition, it is
difficult to know a priori which environmental variables or community attributes
to measure. Finally, while these approaches can establish a potential magnitude
and direction for these relationships, it is often difficult to identify the
taxa or genes that explain the connection between composition and function.

The other broad category of approaches used to address this question are
manipulative approaches. Manipulative experiments try to alter microbial
community composition and observe the effect on function. For example,
reciprocal transplant and common garden experiments have shown that microbiomes
originating from different ecosystems inoculated into the same substrate or
introduced into a common environment display distinct functional rates 
[@cavigelli2000; @balser2005; @waldrop2006; @strickland2009]. In addition,
manipulating diversity by filtering communities by cell size or through dilution
has been shown to alter the rate of ecosystem functions 
[@wagg2014; @wagg2019; @griffiths2001]. However, manipulating the microbiome
directly is challenging. Each of the manipulative approaches applied confound
community composition with other factors. For example, reciprocal transplant and
common garden experiments confound community composition with the abiotic
conditions carried on the inoculum while manipulating composition through
dilution may confound composition with biomass [@reed2007].

(Side note: papers like the Strickland paper were actually asking a subtly
different question from the one that we are asking. Strickland was asking
whether communites can be functionally dissimilar as opposed to functionally
redundant. That is, that historical contingency plays a role in the functioning
of microbial communities. They demonstrated quite well that communities
originating from different environments could have unique functional rates.
However, is that due to microbiome-environment covariance where a different
environment (i.e., a different litter substrate) selects for a litter-degrading
community or is that a result of the unique effect of variation in the
community? They actually point this out in their Discussion, they suggest future
research should determine whether the differences between communities are driven
by taxonomic composition or is function the result of the selection environment
(I think of this is analogous to our "covariation" argument)? They also suggest
trying to adapt communities to a new environment through continuous inputs of
the same resource (just like in our experiment!))

In this study, we sought to build on the observations of comparative and
manipulative studies by applying a different approach to the question of whether
variation in the microbiome contributes to variation in the rate of an ecosystem
function. We used artificial ecosystem selection to select for microbiomes that
performed a greater rate of ecosystem function
[@swenson2000; @blouin2015; @panke-buisse2015]. We then tested whether variation
in the microbiome contributed to variation in the rate of ecosystem function and
identified microbiome attributes that might explain this relationship. There are
several potential advantages to this approach for documenting
microbiome-function relationships and for investigating potential mechanisms
underlying those relationships. By passaging microbiomes over multiple rounds
using a common environment, we can weaken the covariance between microbes and
the environment by repeatedly diluting abiotic effects. In addition, by
inoculating our ecosystems with equally sized subsamples, we eliminate the need
to generate microbiome variation through methods that are confounded with
biomass or cell size. Lastly, by comparing our artificially selected community
to a control community resulting from random selection, we can both control for
changes in the environment over time and identify significant genes or taxa
associated with our selection treatment to establish a potential mechanism that
explains this variation.

We applied artificial ecosystem selection to soil microbiomes by selecting on
soil methane oxidation rate. We chose this function because methane is a
globally important greenhouse gas and methane oxidation by soil bacteria is the
primary biological sink for atmospheric methane as well as possibly the main driver
of temporal methane dynamics [@kirschke2013]. In addition, there is evidence that
soil methane oxidation rate may vary with microbiome composition based on
comparative studies in a variety of arctic and tropical ecosystems 
[@meyer2020; @meyer2020a; @freitag2009; @freitag2010] as well as studies that
manipulate methanotroph richness [@schnyder2018]. Finally, methanotrophy is one
of the most deeply conserved microbial physiologies and is represented in a
narrow range of taxa, which suggests that the taxonomic composition of the
microbiome is more likely to be associated with the rate of methane oxidation
than other broader or more shallowly conserved functions
[@schimel1998; @martiny2013].

In this study, we evaluated whether variation in the microbiome contributes to
variation in the rate of ecosystem functions. To address this problem, we
applied artificial ecosystem selection to methane oxidation rate of the soil
microbiome and answered the following questions: Does variation in the relative
abundance of microbial taxa contribute to variation in soil methane oxidation
rate in the lab? How much of the variation in soil methane oxidation rate can
be attributed to the microbiome? What attributes of the microbiome regulate
variation in methane oxidation rate? And do these attributes match our
assumptions about what regulates methane oxidation rate in nature?

@@

We performed an artificial ecosystem selection experiment 
[*sensu* @swenson2000] by passaging replicate soil microbiomes. The trait we
selected on was CH~4~ oxidation rate. This experiment had two selection lines
with twelve jars each for a total of twenty-four jars per passage. One line was
a positive selection line where the two or three jars with the highest CH~4~
oxidation rate were homogenized to inoculate the next set of positive jars. The
other line was a neutral selection line where an equal number of jars as the
positive line were chosen at random to inoculate the next set of neutral jars.
The number of jars chosen was based on the distribution of fluxes among the
positive jars, i.e., we chose the top three jars unless only two jars had
considerably greater CH~4~ oxidation rates based on visual inspection of
histograms. The experiment was carried out over five passages until a
significant divergence in functional rates was observed between the two
selection treatments based on a difference of slopes.

The initial soil microbiome was sampled from the top 10 cm of an upland mineral
soil under a deciduous forest ecosystem near the University of Oregon campus in
Eugene, OR, USA. Incubations were performed in 500 mL mason jars with rubber
septa installed in the lids. Each jar was sterilized with 70\% ethanol to which
was added 45 g of autoclaved potting mix, 5 g of living soil, and 3.5 mL of
sterile deionized water to bring them to 60\% of field capacity. The potting
mix was Lane Potting Mix from Lane Forest Products (Eugene, OR, USA). Each jar
was then capped and injected with 4.3 mL of 99\% CH~4~ to bring
the headspace concentration to approximately 1000 ppm CH~4~. To create the two
treatment groups, twenty-four jars were created in an identical manner and then
randomly assigned to either the positive or neutral selection line. Jars were
flushed and respiked with methane twice per week to maintain aerobic conditions and elevated
CH~4~ concentrations and were incubated at ambient temperature for
approximately four weeks. Methane oxidation rates were determined at the end
of the incubation period. For the positive treatment, the three jars with the
greatest CH~4~ oxidation rates were chosen to inoculate the next generation.
For the neutral treatment, three jars were randomly selected to inoculate the
next generation. For each treatment, the selected jars were homogenized and 5 g
of the homogenized soil was used to inoculate the next set of jars, which
represents a 10% subsampling. The next set of jars were created in an identical
manner to the first generation with fresh autoclaved potting mix and the same
moisture and CH~4~ content.

@@

We estimated the proportion of variation in methane oxidation rate due to
variation in the microbiome as the regression of divergence between the positive
and neutral treatment on the cumulative selection differential [@falconer1996].
Divergence was calculated as the mean methane oxidation rate of the positive
treatment minus the mean methane oxidation rate of the neutral treatment in each
passage. The selection differential was calculated as the difference between
the mean of the three selected jars and the mean of all twelve jars in a passage.
Cumulative selection differential was then the sum of the selection differential
from all preceding selection events. We test this as the regression of
divergence on cumulative selection differential using the 'lm' function.

Richness was estimated using the method from @hurlbert1971 with a subsample
size of `r min_seqs` using the 'rarefy' function in 'vegan' [@oksanen2019]. We
tested a difference in richness by both passage and treatment with a
Kruskal-Wallace test followed by a pairwise Wilcoxon test. Next, we estimated
beta-diversity as the Bray-Curtis dissimilarity by averaging 100 random subsets
with a subsample size of `r min_seqs` using the 'avgdist' function in 'vegan'
[@bray1957; @oksanen2019].  We tested a difference in centroid and dispersion
of beta diversity by passage and treatment using a permutational analysis of
variance (PERMANOVA) with 999 permutations using the 'adonis2' function from
'vegan' and tested a difference of group dispersions using 'betadisper' and
'anova' with 999 permutations [@mcardle2001; @oksanen2019].  Lastly, we tested
the correlation between CH~4~ oxidation rate and Bray-Curtis dissimilarity in
Passage 5 with a distance-based redundancy analysis (dbRDA) using the 'dbrda'
function in 'vegan' and estimated the p-value using a permutation F-test with
999 permutations [@mcardle2001; @oksanen2019]

@@

We estimated the proportion of variation in methane oxidation rate due to
variation in the microbiome by regressing the divergence between the positive
and neutral treatments against the cumulative selection differential. We
observed a slope of 0.40, which would indicate that 40\% of the variation is due
to the microbiome, but this was not significant (statistics).

We estimated the ecosystem heritability defined as the proportion of variation
in methane oxidation due rate to variation in the microbiome as the regression
of mid-recipient on mid-donor (`r figures("herit", display = "cite")`).
Recipient CH~4~ oxidation rates were correlated with donor CH~4~ oxidation rates
in both the Positive treatment
`r paste0('(', 'slope', ' = ', myround(tidy(herit_model)[2, 2] + tidy(herit_model)[4, 2], 2), ', SE = ', myround(tidy(herit_model)[4, 3], 2), ', t = ', myround(tidy(herit_model)[4, 4], 2), ', p = ', myround(tidy(herit_model)[4, 5], 2), ')')` and 
the Neutral treatment `r print_lm(herit_model, 2)`.

@@

We did not identify any methanotrophic families in the overall differential abundance analysis. 
However, we wanted to look more closely at the known methanotrophs in our dataset to be sure they did not have an effect.
To do this, we subset all of the ASVs in our dataset that were in families that contained methanotrophs.
Only two families were represented: *Methylacidiphilaceae* and *Beijerinckiaceae*. 
Aggregating reads at the family level, neither family was differentially abundant between the two treatments.
However, aggregated at the genus level, a group of unclassified genera in the *Beijerinckiaceae* were depleted in the Positive treatment and the genus
*Rhodoblastus*, a member of the *Beijerinckiaceae*, was enriched in the
Positive treatment.  While many *Beijerinckiaceae* are methanotrophs, several
taxa in this family have lost the the ability to oxidize CH~4~ and it appears
that *Rhodoblastus* species are not able to grow on CH~4~, though they can grow
on methanol [@tamas2014, @dedysh2016]. Based on this analysis,
it appears that no methanotrophs were enriched in the Positive treatment.

@@

In this study, we addressed whether variation in the microbiome contributes to
variation in the rate of ecosystem function. To test this, we applied artificial
ecosystem selection to soil microbiomes by selecting on the whole-ecosystem soil
methane oxidation rate. We observed an increase in methane oxidation rate in the
selection treatment relative to the control, which demonstrates that there was a
response to selection. Given that we observed a response to selection, we
conclude that variation in the microbiome contributes to variation in the
methane oxidation rate independent of the environment.

The ultimate goal of this study (and much of microbial ecology) could be
summarized as "Do microbes matter to ecosystem function (or health)?" and there
are at least three ways we could address that question. One would be to ask
whether microorganisms mediate function. This question is perhaps trivial. In
the case of methane emissions from soil, we know that microorganisms produce and
consume methane. We could also ask what would happen if we removed the
microorganisms. Again, this question is trivial. There are no known abiotic
reactions in soil that consume methane (outside of the troposphere, the
stratosphere, and the marine boundary layer @kirschke2013). Therefore, removing
the microbiome would likely eliminate methane production and consumption. Then,
the third question we could ask is whether *variation* in the microbiome is
associated with *variation* in the rate of ecosystem function. The answer to
this question is not so obvious.

If we look at variation in the rate of a microbially-mediated ecosystem function
across a set of habitats, that variation could be associated with variation in
the abiotic environmental conditions, variation in the composition of
microbiomes, or some interaction or covariance between the two (see Figure 5).
To address the question of whether variation in the microbiome matters for
ecosystem function, we need to identify the component of variation uniquely
associated with the microbiome. If we sample natural variation, as in the
comparative approach, it can be difficult to distinguish between the direct
effect of the microbiome and the effect of the microbiome due to covariance with
the environment. Likewise, with the manipulative approach even though we are
placing ecosystem propagules into a common environment or substrate, the
propagule itself carries with it environmental conditions that
cannot be removed. Generating variation in microbiome composition through
filtering or other similar means often confounds composition with other factors,
such as biomass or cell size. To overcome these limitations and to isolate
the direct effect of the microbiome on ecosystem function, we used
artificial ecosystem selection.

To perform artificial ecosystem selection, we generate a set of ecosystems in
bottles using a common, sterile substrate and inoculate them with a propagule
containing a diverse microbiome that is a subsample of an existing microbiome.
This generates variation in microbiome composition among the ecosystems. We can
then measure the rate of an ecosystem function for each bottle and select the
bottles with the highest (or lowest) rate of ecosystem function. The selected
ecosystems are then homogenized and used as the propagule to inoculate a new set of
ecosystems containing the same sterile substrate as the initial set. By
subsampling the microbiome, we can generate variation in the composition of the
microbiome that is independent of cell size or biomass. And by passaging the
microbiome through the same sterile substrate over multiple rounds of selection,
we can progressively minimize the effect of the abiotic conditions from the
donor environment on both the rate of function and the composition of the
microbiome. In this way, over multiple rounds of selection, we can observe
whether varying the microbiome can generate variation in the rate of our
ecosystem function. If we observe no change in ecosystem function from
passage to passage despite our imposed selection, we would conclude that this
ecosystem function in this set of ecosystems is driven primarily by
environmental variation or technical error among the replicates. However, if we
observe a response to selection, that is, a shift in the mean ecosystem function
between successive passages, we would conclude that variation in the microbiome can
generate variation in the rate of ecosystem function. And that is, in fact, what
we observed.

>>>>

{title max characters 150; current = 106 w/spaces}

@@

```{r initialize-packages}
library(tidyverse)
library(broom)
library(broman)
library(morris)
# library(phyloseq)
library(ggpubr)
library(knitr)
library(corncob)
# library(ggthemes)
library(captioner)
library(knitr)
library(ggtext)
library(patchwork)
figures <- captioner()
tables <- captioner(prefix = "Table")

@@

load(file = paste0(der_dir, 'plots.RData'))
load(file = paste0(der_dir, 'models.RData'))

# # CH4 models
# load(file = paste0(der_dir, 'response_model.RData'))
# load(file = paste0(der_dir, 'response_plot.RData'))
# load(file = paste0(der_dir, 'herit_model.RData'))
# load(file = paste0(der_dir, 'herit_plot.RData'))
# 
# # 16S models
# load(file = paste0(der_dir, 'richness_models.RData'))
# load(file = paste0(der_dir, 'phyloseq.RData'))
# load(file = paste0(der_dir, 'beta_model.RData'))
# load(file = paste0(der_dir, 'beta_fig.RData'))
# load(file = paste0(der_dir, 'da_corncob.RData'))
# load(file = paste0(der_dir, 'da_fig.RData'))
# load(file = paste0(der_dir, 'dbrda.RData'))

@@

{max 250 words; current = 249}

Microbiomes mediate important ecosystem functions, yet it has proven difficult
to determine the relationship between microbiome composition and the rate of
ecosystem functions. This challenge remains because it is difficult to
manipulate microbiome composition directly, we often cannot know _a priori_ which
microbiome members influence the rate of an ecosystem function, and
microbiomes can covary strongly with other drivers of ecosystem function, such
as the environment. We propose that artificial ecosystem selection can overcome
these challenges. Artificial ecosystem selection involves creating replicate
ecosystems, selecting replicates with a desired ecosystem-level trait, and
using the microbiomes from the selected ecosystems to inoculate a new set of
ecosystems. We performed artificial ecosystem selection on replicate soil
microbiomes to select for increased soil methane oxidation rate. We observed a
strong response to selection with a `r myround(per_slope, 1)`\% increase in
methane oxidation rate per passage. We
estimated that 40\% of the variation in methane oxidation
rate in these soils can be attributed to microbiome variation. We also found
that selection did not enrich for known methane oxidizers; instead, `r length(da_fams)` families not
known to oxidize methane, including *Fimbriimonadaceae*, *Cytophagaceae*, and
*Diplorickettsiaceae*, were enriched by selection. This result is in contrast to the typical assumption that
the rate of an ecosystem function is limited by the final step in the associated
microbial pathway. Our study demonstrates that variation in microbiome
composition can contribute to variation in the rate of ecosystem function and
that manipulating microbiome composition could be a viable strategy for managing
ecosystem functions.

{max text 5000 words; current text = 4549}

@@

Microbiomes mediate a variety of important ecosystem functions relevant to human
health, agriculture, and global change. As a result, there is great interest in
understanding how to manipulate the microbiome to achieve desirable outcomes
within these domains [@conrad1996; @schimel1998; @crowther2019]. However, for
microbiome manipulations to be successful, variation in the microbiome must
contribute directly to variation in the magnitude of the function of interest
independent of other factors. Many studies have attempted to document such a
relationship
[@cavigelli2000; @balser2005; @waldrop2006; @strickland2009; @wagg2014; @wagg2019].
However, it is difficult to isolate the direct effect of variation in microbiome
composition from other drivers of variation in ecosystem function, such as the
indirect effect of the environment on function through microbiome assembly.
Here, we overcome these limitations by using a selection approach to estimate
the degree to which an ecosystem function varies with microbiome composition.

There have been two general categories of approaches that investigators have
used to estimate the degree to which an ecosystem function varies with
microbiome composition: comparative and manipulative. Comparative
studies sample natural variation in an ecosystem function across different
habitats and simultaneously measure variation in community composition.
Investigators can then correlate ecosystem function with aspects of community
composition while attempting to control for environmental variation. These
approaches have
documented important relationships between microbiomes and ecosystem
functions. For example, a meta-analysis of these studies observed a small but
significant contribution of the microbiome to variation in ecosystem
function after controlling for environmental variation [@graham2016]. In
addition, studies focusing on the correlation between the rate of an ecosystem
function and the abundance of an associated marker gene (i.e., a gene that codes
for a protein assumed to be involved in the ecosystem function) sometimes observe a
significant correlation, though this relationship is rare and contingent upon
both the function and the ecosystem sampled [@rocca2015]. However, comparative
studies come with some unique challenges and limitations. One issue is that
microbiome attributes tend to covary with the abiotic conditions within an
environment, and whether and how the investigator controls for these variables
can affect the conclusions of such a comparative study. In addition, it is
difficult to know _a priori_ which environmental variables or community attributes
to measure. Finally, while these approaches can establish a potential magnitude
and direction for these relationships, it is often difficult to identify the
taxa or genes that explain the connection between composition and function.

The other broad category of approaches used to address this question are
manipulative approaches. Manipulative experiments try to alter microbial
community composition and observe the effect on function. For example,
reciprocal transplant and common garden experiments have shown that microbiomes
originating from different ecosystems inoculated into the same substrate or
introduced into a common environment display distinct functional rates 
[@cavigelli2000; @balser2005; @waldrop2006; @strickland2009]. In addition,
manipulating diversity by filtering communities by cell size or through dilution
has been shown to alter the rate of ecosystem functions 
[@wagg2014; @wagg2019; @griffiths2001]. However, manipulating the microbiome
directly is challenging. Each of the manipulative approaches applied confound
community composition with other factors. For example, reciprocal transplant and
common garden experiments can confound community composition with the abiotic
conditions introduced with the inoculum, while manipulating composition through
dilution may confound composition with biomass [@reed2007].

In this study, we sought to build on the observations of comparative and
manipulative studies by applying a different approach to the question of whether
microbiome variation contributes to variation in the rate of an ecosystem
function. We used artificial ecosystem selection to select for microbiomes that
performed a greater rate of ecosystem function
[@swenson2000; @blouin2015; @panke-buisse2015]. We then tested whether variation
in the microbiome contributed to variation in the rate of ecosystem function and
identified microbiome attributes that might explain this relationship. There are
several potential advantages to this approach for documenting
microbiome-function relationships and for investigating the mechanisms
underlying those relationships. By passaging microbiomes over multiple rounds
using a common environment, we can weaken the covariance between microbes and
the environment by repeatedly diluting abiotic effects. In addition, by
inoculating our ecosystems with equally-sized subsamples, we eliminate the need
to generate microbiome variation through methods that are confounded with
biomass or cell size. Lastly, by comparing our artificially selected community
to a control community resulting from random selection, we can both control for
changes in the environment over time and identify genes or taxa that are
associated with the ecosystem function under selection.

We applied artificial ecosystem selection to soil microbiomes by selecting on
soil methane oxidation rate. We chose this function because methane is a
globally important greenhouse gas and methane oxidation by soil bacteria is the
primary biological sink for atmospheric methane [@kirschke2013]. In addition,
there is evidence that soil methane oxidation rate may vary with microbiome
composition based on comparative studies in a variety of arctic and tropical
ecosystems [@meyer2020; @meyer2020a; @freitag2009; @freitag2010] as well as
studies that manipulate methanotroph richness [@schnyder2018]. Finally,
methanotrophy is one of the most deeply conserved microbial physiologies and is
represented in a narrow range of taxa, which suggests that the taxonomic
composition of the microbiome is more likely to be associated with the rate of
methane oxidation than other broader or more shallowly conserved functions
[@schimel1998; @martiny2013].

In this study, we used artificial ecosystem selection on methane oxidation rate
to address the following questions: Does variation in the relative abundance of
microbial taxa contribute to variation in soil methane oxidation rate in our
system? Which attributes of the microbiome are associated with variation in
methane oxidation rate, and do these attributes match our assumptions about the
factors that regulate methane oxidation rate in nature?

@@

We performed an artificial ecosystem selection experiment 
[*sensu* @swenson2000] by passaging replicate soil microbiomes. The trait we
selected on was methane (CH~4~)-oxidation rate. This experiment had two selection lines
with twelve jars each for a total of twenty-four jars per passage. One line was
a positive selection line where the two or three jars with the highest CH~4~
oxidation rate were homogenized to inoculate the next set of positive jars. The
other line was a neutral selection line where an equal number of jars as the
positive line were chosen at random to inoculate the next set of neutral jars.
The number of jars chosen was based on the distribution of fluxes among the
positive jars, i.e., we chose the top three jars unless only two jars had
considerably greater CH~4~ oxidation rates based on visual inspection of
histograms. The experiment was carried out over five passages until a
significant divergence in functional rates was observed between the two
selection treatments based on a difference of slopes.

The initial soil microbiome was sampled from the top 10 cm of an upland mineral
soil under a deciduous forest ecosystem near the University of Oregon campus in
Eugene, OR, USA. Incubations were performed in 500 mL mason jars with rubber
septa installed in the lids. Each jar was sterilized with 70\% ethanol to which
was added 45 g of autoclaved potting mix, 5 g of living soil, and 3.5 mL of
sterile deionized water to bring them to 60\% of field capacity. The potting
mix was Lane Potting Mix from Lane Forest Products (Eugene, OR, USA). Each jar
was then capped and injected with 4.3 mL of 99\% CH~4~ to bring
the headspace concentration to approximately 1000 ppm CH~4~. To create the two
treatment groups, twenty-four jars were created in an identical manner and then
randomly assigned to either the positive or neutral selection line. Jars were
flushed and respiked with CH~4~ twice per week to maintain aerobic conditions
and elevated CH~4~ concentrations and were incubated at ambient temperature for
approximately four weeks. Methane oxidation rates were determined at the end of
the incubation period. For the positive treatment, the three jars with the
greatest CH~4~ oxidation rates were chosen to inoculate the next generation.
For the neutral treatment, three jars were randomly selected to inoculate the
next generation. For each treatment, the selected jars were homogenized and 5 g
of the homogenized soil was used to inoculate the next set of jars, which
represents a 10% subsampling. The next set of jars were created in an identical
manner to the first generation with fresh autoclaved potting mix and the same
moisture and CH~4~ content.

@@

We estimated the proportion of variation in methane oxidation rate due to
variation in the microbiome as the regression of divergence between the positive
and neutral treatment on the cumulative selection differential [@falconer1996].
Divergence was calculated as the mean methane oxidation rate of the positive
treatment minus the mean methane oxidation rate of the neutral treatment in each
passage. The selection differential was calculated as the difference between
the mean of the three selected jars and the mean of all twelve jars in a passage.
Cumulative selection differential was calculated as the sum of the selection
differential from all preceding selection events. We then regressed cumulative
divergence on cumulative selection differential using the 'lm' function.

Richness was estimated using the method from @hurlbert1971 with a subsample
size of `r min_seqs` calculated via the 'rarefy' function in 'vegan' [@oksanen2019]. We
tested a difference in richness by both passage and treatment with a
Kruskal-Wallace test followed by a pairwise Wilcoxon test. Next, we estimated
beta-diversity as the Bray-Curtis dissimilarity by averaging 100 random subsets
with a subsample size of `r min_seqs` using the 'avgdist' function in 'vegan'
[@bray1957; @oksanen2019].  We tested a difference in centroid and dispersion
of beta diversity by passage and treatment using a permutational analysis of
variance (PERMANOVA) with 999 permutations using the 'adonis2' function from
'vegan' and tested a difference of group dispersions using 'betadisper' and
'anova' with 999 permutations [@mcardle2001; @oksanen2019].  Lastly, we tested
the correlation between CH~4~ oxidation rate and Bray-Curtis dissimilarity in
Passage 5 with a distance-based redundancy analysis (dbRDA) using the 'dbrda'
function in 'vegan' and estimated the p-value using a permutation F-test with
999 permutations [@mcardle2001; @oksanen2019]

@@

To estimate the proportion of variation in CH~4~ oxidation rate due to
variation in microbiome composition, we regressed divergence between the
positive selection treatment and the control against the cumulative selection
differential (`r figures("herit", display = "cite")`). This estimate is analogous
to estimates of "microbiability" from the animal breeding literature, which
quantifies the
variation in a host trait that is due to microbiome variation [@difford2016]. 
The slope of the regression of divergence on cumulative selection differential
provides an estimate of realized microbiability (h^2^ &pm; SE), which in our
experiment was `r myround(tidy(rh_model)[[2, 2]], 2)` &pm;
`r myround(tidy(rh_model)[[2, 3]], 2)`, though this was not significant (F~`r glance(rh_model)$df`,`r glance(rh_model)$df.residual`~ = 
`r myround(glance(rh_model)$statistic, 2)`, 
p = `r myround(glance(rh_model)$p.value, 2)`).

@@

We did not identify any methanotrophic families in the overall differential abundance analysis. 
However, we wanted to look more closely at the known methanotrophs in our dataset to be sure that they did not have an effect.
To do this, we subset the ASVs in our dataset that were in families that contained methanotrophs.
Only two families were represented: *Methylacidiphilaceae* and *Beijerinckiaceae*. 
Aggregating reads at the family level, neither family was differentially abundant between the two treatments.
However, aggregated at the genus level, a group of unclassified genera in the *Beijerinckiaceae* were depleted in the Positive treatment and the genus
*Rhodoblastus*, a member of the *Beijerinckiaceae*, was enriched in the
Positive treatment.  While many *Beijerinckiaceae* are methanotrophs, several
taxa in this family have lost the ability to oxidize CH~4~ and it appears
that *Rhodoblastus* species are not able to grow on CH~4~, though they can grow
on methanol [@tamas2014; @dedysh2016]. Based on this analysis,
it appears that no methanotrophs were enriched in the Positive treatment.

@@

We used artificial ecosystem selection to estimate the contribution of variation
in microbiome composition to variation in the rate of an ecosystem function,
methane oxidation in soil, independent of environmental variation. Understanding
how and to what degree microbiome variation contributes to variation in
ecosystem function is important for many reasons. For example, successful
microbiome manipulations require that the manipulated microbiome contribute to
variation in ecosystem function independent of other drivers of ecosystem
variation (such as variation in environmental conditions). This is because the
drivers of variation in ecosystem functions can interact in complicated ways
(Figure 5);
for example, environmental variation can directly contribute by providing
conditions that select for microbial groups that in turn alter the rates of
ecosystem functions. Determining the independent contribution of microbiome
variation to ecosystem function is crucial because if microbiome composition is
driven primarily by environmental conditions, then introducing a desirable taxon
through microbiome manipulation without altering the environment will likely be
unsuccessful at shifting the targeted ecosystem function. 
The artificial selection approach is different from the comparative and
manipulative approaches used in past attempts at answering this question,
because it can control for both the direct effect of environment on function as
well as the indirect effect on environment via its impact on microbiome
assembly.

In our study, we observed an increase in methane oxidation rate in the
selection treatment relative to the control, which demonstrates that there was a
response to selection. Given that we observed a response to selection, we
conclude that variation in the microbiome contributes to variation in the
methane oxidation rate independent of the environment. This suggests that
microbiome manipulations could be an effective approach for altering the rate of
methane oxidation in this soil, and that the artificial selection approach may
be useful in determining the potential for microbiome manipulations for other
functions in other ecosystems.

Given that variation in the microbiome is associated with variation in the rate
of an ecosystem function in our system, a reasonable follow-up question is "how
much variation in ecosystem function is associated with microbiome variation in
this system?" One way to estimate this is to determine how much the recipient
jars resemble the selected donor jars that were used to inoculate them
[@falconer1996]. We can
calculate the response to selection as the difference between two successive
passages in their mean methane oxidation rate. We will denote this as $R$. We
can also calculate the strength of selection as the difference in mean methane
oxidation rate between the twelve jars in one generation and the three jars chosen
for selection in that generation, which we will call the selection differential
and denote as $S$. If we plot the cumulative $R$ against the cumulative $S$, the
slope of this relationship will equal the proportion of variation explained by
the microbiome. If the change in mean function from passage one to passage two
($R$) is equal to the difference in mean function between the twelve jars in
passage one and the three jars selected to inoculate passage two ($S$), then we
would conclude that 100% of the variation is due to variation in the microbiome.
Likewise, if recipients do not resemble the donors in their mean methane
oxidation rate and simply wander randomly, then we would conclude that all of
the variation is due to the environment or technical variation.

This relationship between microbiome variation and ecosystem function variation
is analogous to the concept of "heritability" [@lynch1998] used by quantitative
geneticists, or more precisely the concept of "microbiability" [@difford2016] proposed
by microbiome scientists who study host-associated microbiomes. Although rarely
used in the study of environmental microbiomes, this concept could be very
useful for understanding and manipulating microbially-mediated functions in a
variety of ecosystems. In our experiment, variation in microbiome taxonomic
composition statistically explained (i.e., was associated with) 40% of the
variation we observed in the rate of methane oxidation.  This suggests that
there is substantial potential for altering this ecosystem function through
microbiome manipulation in this soil. It is very likely that the
“environmental microbiability” will be different for other ecosystem functions
in this soil and for methane oxidation in other soils. However, our experiment
demonstrates that this relationship is measurable and provides an example of how
this can be accomplished.

Given that variation in the microbiome is driving substantial variation in
methane oxidation rate in our system, we next wanted to determine which aspects of the
microbiome best explain this contribution of microbiome variation to
methane oxidation rate. There are three inter-related ways that microbiomes could have responded
to selection in this experiment: gain or loss of taxa, changes in the
relative abundances of taxa, or changes within the genomes of the constituent
taxa. We surveyed microbiome variation via 16s rRNA ribotyping in our
experiment, which allowed us to deeply sample taxonomic diversity but did not
allow us to directly address whether taxa in this experiment evolved genomic changes
as a result of selection. However, if such genomic changes resulted in increased
persistence or abundance of the population with these changes, this would be
detectable. Therefore, we will focus on the first
two possibilities.
